---
title: "FAERS-Pharmacovigilance"
output: 
  rmarkdown::html_vignette: default
  github_document: default
vignette: >
  %\VignetteIndexEntry{FAERS-Pharmacovigilance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_format = "github_document",
      output_file = "README.md",
      output_dir = getwd()
    )
  })
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

You can install the development version of `faers` from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
if (!requireNamespace("pak")) {
  install.packages("pak",
    repos = sprintf(
      "https://r-lib.github.io/p/pak/devel/%s/%s/%s",
      .Platform$pkgType, R.Version()$os, R.Version()$arch
    )
  )
}
pak::pkg_install("Yunuuuu/faers")
```

## Pharmacovigilance Analysis using FAERS

FAERS is a database for the spontaneous reporting of adverse events and
medication errors involving human drugs and therapeutic biological products.
This package accelarate the process of Pharmacovigilance Analysis using FAERS.

```{r setup}
library(faers)
```

### check metadata of FAERS
This will return a data.table reporting years, period, quarter, and file urls
and file sizes. By default, this will use the cached file in
`rappdirs::user_cache_dir("faers")`.  If it doesn't exist, the internal will
parse metadata in <`r sprintf("%s/extensions/FPD-QDE-FAERS/FPD-QDE-FAERS.html",
faers:::fda_host("fis"))`> 
```{r}
faers_meta()
```

An metadata copy was associated with the package, just set `internal = TRUE`.
```{r}
faers_meta(internal = TRUE)
```

#### download and parse quarterly data files from FAERS
The FAERS Quarterly Data files contain raw data extracted from the AERS database
for the indicated time ranges. The quarterly data files, which are available in ASCII or SGML formats, include: 

- demographic and administrative information (demo);
- drug information from the case reports (drug);
- reaction information from the reports (reac);
- patient outcome information from the reports (outc);
- information on the source of the reports (rpsr); 
- drug therapy start dates and end dates for the reported drugs (ther); 
- Medical Dictionary for Regulatory Activities (indi)

Generally, we can use `faers()` function to download and parse all quarterly
data files from FAERS. Internally, the `faers()` function seamlessly utilizes
`faers_download()` and `faers_parse()` to preprocess each quarterly data file
from the FAERS repository. The
default `format` was `ascii` and will return a `FAERSascii` object. (xml format
would also be okay , but presently, the XML file receives only minimal support
in the following process.)

```{r}
# # you must change `dir`, as the file included in the package is sampled
data1 <- faers(2004, "q1",
  dir = system.file("extdata", package = "faers"),
  compress_dir = tempdir()
)
data1
```

Furthermore, in cases where multiple quarterly data files are requisite, the
`faers_combine()` function is judiciously employed. 
```{r}
data2 <- faers(c(2004, 2017), c("q1", "q2"),
  dir = system.file("extdata", package = "faers"),
  compress_dir = tempdir()
)
data2
```

#### standardize and De-duplication
The `reac` file provides the adverse drug reactions, where it includes the
“P.T.” field or the “Preferred Term” level terminology from the Medical
Dictionary for Regulatory Activities (MedDRA). The `indi` file contains the drug
indications, which also uses the “P.T.” level of MedDRA as a descriptor for the
drug indication. In this way, `MedDRA` was necessary to standardize this field
and add additional informations, such as `System Organ Classes`. 

One limitation of FAERS database is Duplicate and incomplete reports. There are
many instances of duplicative reports and some reports do not contain all the
necessary information. We deemed two cases to be identical if they exhibited a
full concordance across drugs administered, and adverse reactions and but showed
discrepancies in one or none of the following fields: gender, age, reporting
country, event date, start date, and drug indications. 

```{r, eval=FALSE}
# you should replace `meddra_path` with the directory of meddra data
data <- faers_standardize(data2, meddra_path)
data <- faers_dedup(data)
```

#### Pharmacovigilance analysis
Pharmacovigilance is the science and activities relating to the detection,
assessment, understanding and prevention of adverse effects or any other
medicine/vaccine related problem. 

```{r, eval=FALSE}
faers_phv_signal(data,
  filter_params = list(field = "demo", .fn = function(x) {
    sample(x$primaryid, 100L)
  })
)
```

#### sessionInfo
```{r}
sessionInfo()
```
